---
title: "Codebook"
author: "Ksenia Kharitonova"
date: "22 March 2017"
output: html_document
---

# Instructions

The purpose of this project is to demonstrate my ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. This code book describes the variables, the data, and any transformations or work that was performed to clean up the data. 

## Descriptions of the data

One of the most exciting areas in all of data science right now is wearable computing. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A [full description](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) is available at the site where the data was obtained.

The data for the project is available here:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

The R script `run_analysis.R` does the following:

1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement.
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names.
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

## Step 1

Checking if the working directory with the script does not contain already the data files in a folder "./UCI HAR Dataset" or zip file.
If not, we download the zip file and unzip it in the working directory.

```{r}
path0 <- getwd()
fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
zipfile <- "getdata%2Fprojectfiles%2FUCI HAR Dataset.zip"

if(!file.exists("./UCI HAR Dataset")){
        if(!file.exists(zipfile)){
                download.file(fileUrl,method="curl")
                unzip(zipfile)
        }
        else {
                unzip(zipfile)
        }
}
path_to_data <- file.path(path0 , "UCI HAR Dataset")
```

List the files in the target directory.

```{r}
files<-list.files(path_to_data, recursive=TRUE)
files
```

`README.txt` and `features_info.txt` files give the detailed information on the dataset. The files with the data used for the Project are as follows:

* `test/subject_test.txt`
* `test/X_test.txt`
* `test/y_test.txt`
* `train/subject_train.txt`
* `train/X_train.txt`
* `train/y_train.txt`
* `features.txt`
* `activity_labels.txt`

Reading a list of all feature names `features.txt` which corresponds to variable names of observations from `X_train.txt` and `X_test.txt` (see `features_info.txt` for more details).

```{r}
variable_names <- read.table(file.path(path_to_data, "features.txt"),
                             header=FALSE, stringsAsFactors = FALSE)
```

Reading data sets of all observations for train and test sets. Combining train set and test set in a unified set. Removing intermediate data from the memory. 

```{r}
train_dataset <- read.table(file.path(path_to_data, "train" , "X_train.txt" ),  header = FALSE)
test_dataset <- read.table(file.path(path_to_data, "test" , "X_test.txt" ), header = FALSE)
combined_dataset <- rbind(train_dataset, test_dataset)
rm("train_dataset","test_dataset")
```


## Step 2

Naming columns in the united set by the corresponding variable names. 

```{r}
names(combined_dataset) <- variable_names$V2
```

Extracting only the measurements on the mean `mean()` and standard deviation `std()` for each measurement.

```{r}
combined_dataset <- combined_dataset[grep("mean\\(\\)|std\\(\\)", variable_names$V2, value=TRUE)]
```

Adding information about subjects to the combined dataset with the observations. Removing intermediate data from the memory.

```{r}
subjects_train <- read.table(file.path(path_to_data, 
                                        "train", "subject_train.txt"),
                                        header = FALSE)
subjects_test  <- read.table(file.path(path_to_data, 
                                        "test" , "subject_test.txt"),
                                        header = FALSE)
subjects <- rbind(subjects_train,subjects_test)
names(subjects) <- "Subject"
combined_dataset <- cbind(combined_dataset, subjects)
rm("subjects", "subjects_train", "subjects_test", "variable_names")
```

## Step 3

Reading information about activities and names of activities. 
Adding information about activities and their names to the combined dataset with the observations and subjects. Removing intermediate data from the memory.

```{r}
activity_train <- read.table(file.path(path_to_data, 
                                       "train", "y_train.txt"),
                             header = FALSE)
activity_test <-  read.table(file.path(path_to_data, 
                                       "test" , "subject_test.txt"),
                             header = FALSE)

activity <- rbind(activity_train,activity_test)

activity_labels <- read.table(file.path(path_to_data, 
                                "activity_labels.txt"),
                                header = FALSE,
                                stringsAsFactors = FALSE)

names(activity) <- "Activity"

activity$Activity <- factor(activity$Activity, 
       levels = activity_labels[,1], 
       labels = activity_labels[,2])

combined_dataset <- cbind(combined_dataset, activity)

rm("activity", "activity_train", "activity_test", "activity_labels")
```

Let's take a look at the resulting dataset.

```{r}
str(combined_dataset)
```

Apparently, data is presented in the wide form.

## Step 4

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix `t` to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the `f` to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

The set of variables of interest for the project that were estimated from these signals are: 

* `mean()`: Mean value
* `std()`: Standard deviation

Replacing the data set variable names with more descriptive names.

```{r}
names(combined_dataset)<-gsub("^t", "time", names(combined_dataset))
names(combined_dataset)<-gsub("^f", "frequency", names(combined_dataset))
names(combined_dataset)<-gsub("Acc", "Accelerometer", names(combined_dataset))
names(combined_dataset)<-gsub("Gyro", "Gyroscope", names(combined_dataset))
names(combined_dataset)<-gsub("Mag", "Magnitude", names(combined_dataset))
names(combined_dataset)<-gsub("BodyBody", "Body", names(combined_dataset))
```

Let's take a loot at those names.

```{r}
names(combined_dataset)
```

## Step 5

From the data set in step 4, creating a second, independent tidy data set with the average of each variable for each activity and each subject.

```{r}
library(plyr)
tidy_dataset <- aggregate(. ~ Subject + Activity, combined_dataset, mean)
tidy_dataset <- tidy_dataset[order(tidy_dataset$Subject,tidy_dataset$Activity),]
```

Writing the resulting dataset to a file `tidy_dataset.txt`.

```{r}
write.table(tidy_dataset, file = "tidy_dataset.txt",row.name=FALSE)
```

